Staged Abstract Interpreters

### 2019-09-08 paper/20_prelim.tex
space
Typos

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {
    def pure[A](a: A): M[A]
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]
  }
    \end{lstlisting}
    \vspace{-0.5em}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \vspace{-0.5em}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}
\vspace{-1em}

@@

Note that here, the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified version of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
of monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad as an argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the one from Haskell.

@@

In this section, we present the generic interface in the style of a big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type both abstract, so that they can be instantiated
differently; we also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

\paragraph{Basic Types}
We start with some basic type definitions that are used in the interpreter. The
identifiers in the program are represented by strings. To represent states, two
required components for the interpreter are environments @Env@ and stores
@Store@. @Env@ maps identifiers to addresses, and @Store@ maps
addresses to values, respectively. The @Env@ captures bound variables in the
scope of current control, and @Store@ models a persistent heap through the
program run-time.  At this moment, the domains of addresses and values are
still abstract, and hence, they are declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

%\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {
    def map[B](f: R[A] => R[B]): M[B]
    def flatMap[B](f: R[A] => M[B]): M[B]
  }

>>>>

\begin{figure}[h!]
  \centering
  \vspace{-1em}
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {
    def pure[A](a: A): M[A]
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]
  }
    \end{lstlisting}
    \vspace{-0.5em}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \vspace{-0.5em}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}
\vspace{-1em}

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad as an argument and produce another monad. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the one from Haskell.

@@

In this section, we present the generic interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently; we also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

\paragraph{Basic Types} We start with some basic type definitions that are used
in the interpreter. The identifiers in the program are represented by strings.
To represent states, two required components for the interpreter are
environments, represented by type @Env@, and stores, represented by type
@Store@. An @Env@ maps identifiers to addresses, and a @Store@ maps addresses to
values. The @Env@ captures bound variables in the scope of current control, and
@Store@ models a persistent heap through the program run-time.  Currently,
the domains of addresses and values are still abstract; hence, they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

%\vspace{-1em}
\begin{figure}[h!]
  \centering
\vspace{-1em}
  \begin{subfigure}[b]{0.45\textwidth}
    \begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {
    def map[B](f: R[A] => R[B]): M[B]
    def flatMap[B](f: R[A] => M[B]): M[B]
  }
